Ubiquitous Digital Technologies and Spatial Structure; an update

### 2020-12-01 paper/2_resubmission/tranos_ioannides.Rmd
responses

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the Zipf coefficient for a broad sample of countries over time.
Zipf coefficient is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations. Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

This leads to an estimate of $\zeta$, known as the Zipf coefficient, but more appropriately the Pareto exponent, when $\zeta$ is not estimated to be near $1$.
While much of the literature focuses on the Zipf coefficient being close to $1$, several estimations of city size distributions lead to estimates  of $\zeta$ that are not necessarily equal to $1$, or even near it, in which case we refer to $\zeta$ as the Pareto, or power law, exponent. Given that our aim here is to estimate $\zeta$ for a number of countries over time as a measure of dispersion, equation \ref{eq1} describes the rank of city $i$ in country $c$ in year $t$:

@@

zipf.table <- net.model$model %>%
  dplyr::select(`as.factor(country)`, `as.factor(year)`, b) %>%
  spread(`as.factor(year)`, b) %>%
  rename(`Zipf coef.` = `as.factor(country)`)

@@

kable(zipf.table, format = "latex",
      caption = "Zipf coefficients\\label{zipf}",
      booktabs = T,
      digits = 2) %>%
  footnote(general = "Corrected as per Gabaix and Ibragimov (2011)") %>%
  kable_styling(latex_options = "scale_down",
                position = "left")
```

@@

stargazer(as.data.frame(data[,c(5:18)]),
          type = "latex",
          digits = 1,
          header = FALSE,
          no.space=TRUE,
          title = "Descriptive statistics for the multi-country model\\label{desc.global}",
          font.size = "footnotesize",
          column.sep.width = "1pt",
          summary.stat = c("n", "min", "mean", "sd", "max"),
          covariate.labels=c("Zipf coefficient",
                             "Zipf inv. sq. error\\textsuperscript{a}",
                             "Population density",
                             "Government expenditure (\\% GDP)",
                             "Trade (\\% GDP)",
                             "GDP growth",
                             "GDP per capita (log)",
                             "Female labour force (\\%)",
                             "Population",
                             "Non agriculture value added (\\% GDP)",
                             "Internet users per 100 hab.",
                             "Broadband users per 100 hab.",
                             "Mobile phone users per 100 hab.",
                             "Fixed phone users per 100 hab."),
          notes = c("\\textsuperscript{a}This is the inverse squared standard error of the estimated Zipf, coefficient, which is used for weighting", 
                    "the observations for the estimation, of equation 3. See Section 4"))
```

@@

The availability of a panel dataset for city sizes across countries enables us to use country fixed effects, which can address potential endogeneity issues related to unobserved country specific characteristics of city size distributions.
However, such a strategy does not address potential simultaneity issues.
Simply put, internet penetration might be affected by spatial structure, as reflected in Zipf coefficients, or both internet penetration and spatial structure might be jointly determined by a third variable.
E.g., if a country already has a dispersed spatial structure, internet is particularly suitable in facilitating communication.
Potential endogeneity in our specification will prevent us from being able to determine the causal impact of internet and digital communication technologies usage on spatial structure, which is the main aim of this paper. In order to address this problem, we will adopt an instrumented variable strategy.
Table \ref{desc.global} also includes the descriptive statistics for the instrumental variable we are using and ($female\:labour\:force$), will be discussed in [Section 4](#sec4).

@@

This section presents the estimation results of \ref{eq3}.
The LHS variable is $\zeta$, the Zipf coefficient, as estimated according to the Gabaix and Ibragimov [@gabaix_rank_2011] correction.
The main explanatory variables of interest are provided at the top of Table \ref{ols.global}; namely internet, broadband, mobile and fixed telephony per $100$ habitants,  expressed in natural logarithms.
The main variables of interest are introduced successively on their own in the regressions reported in Table \ref{ols.global}.
All regressions include country fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared standard error of the estimated Zipf coefficient (see Table \ref{desc.global}) to address potential noise that is carried over from the first part of our identification strategy.
In regards to the interpretation of the estimated coefficients, given that the Zipf coefficient has entered the regression not as an absolute value, but instead as a real number a positive coefficient for a RHS variable indicates an impact towards the decrease of the spatial dispersion of population. In other words, a positive coefficient indicates an effect towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies because of the expansion of digital technologies.

@@

stargazer(net.model, bb.model, fixed.model, mobile.model, type="latex",
          header = FALSE,
          no.space=TRUE,
          title = "OLS estimation of equation (3)\\label{ols.global}",
          font.size = "small",
          column.sep.width = "1pt",
          se = list(net.rse[,"Std. Error"], bb.rse[,"Std. Error"], fixed.rse[,"Std. Error"], mobile.rse[, "Std. Error"]),
          dep.var.labels=c("Zipf Coefficient 2000-18"),
          column.separate = c(1,1,1,1),
          covariate.labels=c("Internet users per 100 hab. (log)",
                             "Broadband users per 100 hab. (log)",
                             "Mobile phone users per 100 hab. (log)",
                             "Fixed phone users per 100 hab. (log)",
                             "Population density (log)",
                             "Government expenditure (% GDP)",
                             "Trade (\\% of GDP)",
                             "Non agriculture value added (\\% GDP)",
                             "GDP growth",
                             "GDP per capita (log)",
                             "Population (log)"),
          single.row = FALSE, df = FALSE,
          omit = c("country", "year"),
          add.lines = list(c("Country fixed effects", "Yes", "Yes", "Yes", "Yes"),
                           c("Yearly fixed effects", "Yes", "Yes", "Yes", "Yes")),
          omit.stat = c("rsq", "f"),
          notes = "Robust Std. Errors in parenthesis") #, corrected Zipf coefficient following Gabaix and Ibragimov (2011)")
```

Table \ref{ols.global} and \ref{2sls.global} report the estimation results based on the UN urban agglomeration data.
Starting with the former, we note that an agglomerative effect is only detected for the internet users and while a marginally opposing result can be seen for broadband users as indicated by the significant coefficients in columns (1) and (2).
For the telephony variables the estimation of \ref{eq3} did not yield statistically significant coefficients.
Before discussing further these results and the effects of the other control variables, we need to highlight that the main challenge of estimating equation \ref{eq3} is the potential endogenous nature of the share of internet users which might prevent us from being able to infer a truly causal effect.
Endogeneity might be an issue here as spatial structure, which is proxied by the Zipf coefficient, might be affected by another source, which also affects internet penetration.
For instance, economic development might affect the concentration of population in large cities and at the same time enable more people to go online.
If we do not address this issue, the coefficient for the main variable of interest will capture potential effects that internet penetration has on spatial structure, but also potential reverse causality effects that spatial structure might generate on internet penetration.
To overcome this potential problem, Table \ref{2sls.global} reports estimates of equation \ref{eq3} using two-stage least squares (2SLS) with instrumental variables (IVs).
The latter are variables which are correlated with our endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any -- of the internet and digital communications on spatial structure.
At a first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs and the other control variables are used instead of the endogenous variable to estimate eq. \ref{eq3}.
A significant effect will verify the causal impact of the internet and digital communication usage on spatial structure.

@@

stargazer(net.2sls, bb.2sls, fixed.2sls, mobile.2sls, type = "latex",
          se = list(net.2sls.rse[,"Std. Error"], bb.2sls.rse[,"Std. Error"],
                    fixed.2sls.rse[,"Std. Error"], mobile.2sls.rse[,"Std. Error"]),
          header = FALSE,
          no.space=TRUE,
          title = "2SLS estimation of equation (3)\\label{2sls.global}",
          font.size = "small",
          column.sep.width = "1pt",
          dep.var.labels=c("Zipf Coefficient 2000-18"),
          column.separate = c(1,1,1,1),
          covariate.labels=c("Internet users per 100 hab. (log)",
                             "Broadband users per 100 hab. (log)",
                             "Mobile phone users per 100 hab. (log)",
                             "Fixed phone users per 100 hab. (log)",
                             "Population density (log)",
                             "Government expenditure (\\% GDP)",
                             "Trade (\\% of GDP)",
                             "Non agriculture value added (\\% GDP)",
                             "GDP growth",
                             "GDP per capita (log)",
                             "Population (log)"),
          single.row = FALSE, df = FALSE,
          omit = c("country", "year"),
          omit.stat = c("rsq", "f"),
          add.lines = list(c(rownames(summ.net.2sls$diagnostics)[1], # use rownames as line name. in this case Weak insterumetns
                             round(summ.net.2sls$diagnostics[1, "statistic"], 2),
                             round(summ.bb.2sls$diagnostics[1, "statistic"], 2),
                             round(summ.fixed.2sls$diagnostics[1, "statistic"], 2),
                             round(summ.mobile.2sls$diagnostics[1, "statistic"], 2)),
                           c(rownames(summ.net.2sls$diagnostics)[2],
                             round(summ.net.2sls$diagnostics[2, "statistic"], 2),
                             round(summ.bb.2sls$diagnostics[2, "statistic"], 2),
                             round(summ.fixed.2sls$diagnostics[2, "statistic"], 2),
                             round(summ.mobile.2sls$diagnostics[2, "statistic"], 2)),
                           c("P-value",
                             round(summ.net.2sls$diagnostics[2, "p-value"], 2),
                             round(summ.bb.2sls$diagnostics[2, "p-value"], 2),
                             round(summ.fixed.2sls$diagnostics[2, "p-value"], 2),
                             round(summ.mobile.2sls$diagnostics[2, "p-value"], 2)),
                           # c(rownames(summ.net.2sls$diagnostics)[3],
                           #   round(summ.net.2sls$diagnostics[3, "statistic"], 2),
                           #   round(summ.bb.2sls$diagnostics[3, "statistic"], 2),
                           #   round(summ.fixed.2sls$diagnostics[3, "statistic"], 2),
                           #   round(summ.mobile.2sls$diagnostics[3, "statistic"], 2)),
                           # c("P-value",
                           #   round(summ.net.2sls$diagnostics[3, "p-value"], 2),
                           #   round(summ.bb.2sls$diagnostics[3, "p-value"], 2),
                           #   round(summ.fixed.2sls$diagnostics[3, "p-value"], 2),
                           #   round(summ.mobile.2sls$diagnostics[3, "p-value"], 2)),
                           c("Country fixed effects", "Yes", "Yes", "Yes", "Yes"),
                           c("Yearly fixed effects", "Yes", "Yes", "Yes", "Yes")),
          notes = c("Robust Std. Errors in parenthesis",
                    "IV: Female participation in labour force"))
```

The main challenge for such an exercise is to find a valid IV.
In our case, the challenge is even bigger as we need to find a variable which is not only fit for purpose, but it also contain as few missing values as possible in order to retain the same number of observations as the OLS models.
This is particularly difficult for a diverse sample of countries as the one we are dealing with here.
We propose here the percentage of female participation in the labor force, which is directly related to digital infrastructure, but not to spatial structure (see Table \ref{desc.global} for the descriptive statistics).
\color{blue}
The literature provides empirical and theoretical evidence that ICT are correlated with increase in female participation in labour markets mostly because they enable teleworking [@dettling2017; @arntz2019working; @watson2018digitalisation; @Lechman2020; @Valberg2020]
\color{black}
At the same time, female labor participation is not related with spatial structure especially given the diverse sample of countries.
<!-- Firstly, the correlation coefficient between the IV and the Zipf coefficient is very low ($-0.14$). -->
Although one might think that increased female labor participation may lead to relocation of households to large cities and therefore affect the spatial structure of cites, such a process varies a lot within our dataset.
Even if this might be true for developed countries with mature urban systems for which a location within a large urban center might provide opportunities for both male and female workers to find jobs, our dataset also includes countries, $41$ per cent of the GDP of which is attributed to primary activities (see Table \ref{desc.global}).
So, female labor participation might also be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the weak identification first stage F-test exceeds by far all the Stock-Yogo critical values for all, but the mobile telephony regression.
Hence, we are confident to interpret the effects of internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and telephony penetration have further enhanced agglomeration forces, at least for large urban agglomerations, while the opposite happens for fixed telephony.
These results are robust against endogeneity issues, but are limited to urban agglomeration included in our data.
Indeed, for the estimation of the Zipf coefficient we only included urban agglomeration of $300,000$ inhabitants or more due to data availability.
Hence, the above estimations cannot verify whether such an effect is also valid for smaller cities.
Interestingly, the above results regarding the role of fixed telephony are in agreement with previous results from [@ioannides2008effect] [@tranos2020ict] which used a smaller panel of countries but without constraints regarding the city size.
However, our results do not support the findings of [@tranos2020ict] regarding internet usage.
In order to overcome the city size limitation, the next section presents two case studies, for which we have obtained much more granular data and therefore we are in a position to test the effect of the internet on the tail of the urban population distribution for these countries.

@@

stargazer(as.data.frame(data.for.decr),
          type="latex",
          summary = T,
          summary.stat = c("n", "min", "mean", "sd", "max"),
          header = FALSE,
          no.space=TRUE,
          title = "Descriptive statistics for the USA model\\label{desc.usa}",
          font.size = "small",
          single.row = FALSE,
          column.sep.width = "1pt",
          covariate.labels=c("normalized difference in ranks, 2013-18",
                             "households w. internet 2013, \\%",
                             "population 2013",
                             "\\% of unemployment 2013",
                             "\\% of white population 2013",
                             "income 2013",
                             "population density 2013",
                             "employment in service 2011, \\%",
                             "commute in minutes 2013",
                             "pop. above 25 w. Bachelor's \\\\
                             degree 2005, \\%",
                             "commute in minutes, 2005"))
```

The LHS variable we use here is the outcome of equation \ref{eq5} in a normalized form^[$$Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.$$]
, so as that variable is bounded between $0$ and $1$ and with a mean of $0.5$.
This transformation enables us to estimate \ref{eq4} not only with OLS, but also with a quasibinomial GLM estimator given that the original form of our LHS variable does not vary continuously, but is instead defined as a difference between two count variables, which may also assume negative values.
The results of the different estimations remain qualitatively the same regardless of the form of the LHS variable we use.
Table \ref{ols.usa} reports the results of the estimation of equation \ref{eq4} using OLS.
The sign and significance of the main variable of interest -- percentage of population with computer and broadband connection -- verifies our global level results.
That is an increase in internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that are OLS and GLM.
In addition, Table \ref{ols.usa} also reports the estimates of interaction effects between the share of population with broadband connection with population and population density.
Although the interaction terms are not significant, the main effect remains qualitatively unchanged.

@@

data.for.decr <- data %>%
  dplyr::select(r_diff_tr, download2011, pop2011, n.tests2011,  
         unemployment.share, british.pop.share, pop.dens,
         work.home.share, services.share, uni.freq) %>%
  mutate(n.tests2011perhab = n.tests2011/pop2011) %>%
  dplyr::select(-n.tests2011) %>%
  dplyr::select(r_diff_tr, download2011, pop2011, n.tests2011perhab,  # re-redring
         unemployment.share, british.pop.share, pop.dens,
         work.home.share, services.share, uni.freq)

>>>>

\color{blue}
The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue, as it is clear that not all economic activities might continue in an
unencumbered manner; see Dingel and Neiman [@dingel2020many].
\color{black}

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
\color{blue}
Pareto exponent coefficient
\color{black}
for a broad sample of countries over time.
\color{blue}
The Pareto coefficient
\color{black}
is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations^[
\color{blue}
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.].
\color{black}
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

\color{blue}
This leads to an estimate of $\zeta$, known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere to for
the remained of the text, even  when $\zeta$ is estimated to be near $1$.
The term Zipf coefficient should be reserved for the special case of
$\zeta=1$.
While much of the literature focuses on the Pareto exponent being
close to $1$, several estimations of city size distributions lead to estimates
of $\zeta$ that are not necessarily equal to $1$, or even near it, in which
case we refer to $\zeta$ as the Pareto, or power law, exponent.
Given that our aim here is to estimate $\zeta$ for a number of countries over
time as a measure of dispersion, equation \ref{eq1} describes the rank of city
$i$ in country $c$ in year $t$:
\color{black}

@@

zipf.table <- net.model$model %>%
  dplyr::select(`as.factor(country)`, `as.factor(year)`, b) %>%
  spread(`as.factor(year)`, b) %>%
  rename(`Countries` = `as.factor(country)`)

@@

kable(zipf.table, format = "latex",
      caption = "Pareto exponent\\label{zipf}",
      booktabs = T,
      digits = 2) %>%
  footnote(general = "Corrected as per Gabaix and Ibragimov (2011)") %>%
  kable_styling(latex_options = "scale_down",
                position = "left")
```

@@

stargazer(as.data.frame(data[,c(5:18)]),
          type = "latex",
          digits = 1,
          header = FALSE,
          no.space=TRUE,
          title = "Descriptive statistics for the multi-country model\\label{desc.global}",
          font.size = "footnotesize",
          column.sep.width = "1pt",
          summary.stat = c("n", "min", "mean", "sd", "max"),
          covariate.labels=c("Pareto exponent",
                             "Pareto exp. inv. sq. error\\textsuperscript{a}",
                             "Population density",
                             "Government expenditure (\\% GDP)",
                             "Trade (\\% GDP)",
                             "GDP growth",
                             "GDP per capita (log)",
                             "Female labour force (\\%)",
                             "Population",
                             "Non agriculture value added (\\% GDP)",
                             "Internet users per 100 hab.",
                             "Broadband users per 100 hab.",
                             "Mobile phone users per 100 hab.",
                             "Fixed phone users per 100 hab."),
          notes = c("\\textsuperscript{a}This is the inverse squared standard error of the estimated Pareto exponent, which is used for weighting",
                    "the observations for the estimation of equation 3. See Section 4"))
```

@@

The availability of a panel dataset for city sizes across countries enables us to use country fixed effects, which can address potential endogeneity issues related to unobserved country specific characteristics of city size distributions.
However, such a strategy does not address potential simultaneity issues.
Simply put, internet penetration might be affected by spatial structure, as reflected in Pareto exponents, or both internet penetration and spatial structure might be jointly determined by a third variable.
E.g., if a country already has a dispersed spatial structure, internet is particularly suitable in facilitating communication.
Potential endogeneity in our specification will prevent us from being able to determine the causal impact of internet and digital communication technologies usage on spatial structure, which is the main aim of this paper. In order to address this problem, we will adopt an instrumented variable strategy.
Table \ref{desc.global} also includes the descriptive statistics for the instrumental variable we are using and ($female\:labour\:force$), will be discussed in [Section 4](#sec4).

@@

This section presents the estimation results of \ref{eq3}.
The LHS variable is $\zeta$, the
\color{blue}
Pareto exponent,
\color{black}
as estimated according to the Gabaix and Ibragimov [@gabaix_rank_2011] correction.
The main explanatory variables of interest are provided at the top of Table \ref{ols.global}; namely internet, broadband, mobile and fixed telephony per $100$ habitants, expressed in natural logarithms.
The main variables of interest are introduced successively on their own in the regressions reported in Table \ref{ols.global}.
All regressions include country fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared standard error of the estimated
\color{blue}
Pareto exponent
\color{black}
(see Table \ref{desc.global}) to address potential noise that is carried over from the first part of our identification strategy.
In regards to the interpretation of the estimated coefficients, given that the
\color{blue}
Pareto exponent
\color{black}
has entered the regression not as an absolute value, but instead as a real number a positive coefficient for a RHS variable indicates an impact towards the decrease of the spatial dispersion of population. In other words, a positive coefficient indicates an effect towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies because of the expansion of digital technologies.

@@

stargazer(net.model, bb.model, fixed.model, mobile.model, type="latex",
          header = FALSE,
          no.space=TRUE,
          title = "OLS estimation of equation (3)\\label{ols.global}",
          font.size = "small",
          column.sep.width = "1pt",
          se = list(net.rse[,"Std. Error"], bb.rse[,"Std. Error"], fixed.rse[,"Std. Error"], mobile.rse[, "Std. Error"]),
          dep.var.labels=c("Pareto exponent 2000-18"),
          column.separate = c(1,1,1,1),
          covariate.labels=c("Internet users per 100 hab. (log)",
                             "Broadband users per 100 hab. (log)",
                             "Mobile phone users per 100 hab. (log)",
                             "Fixed phone users per 100 hab. (log)",
                             "Population density (log)",
                             "Government expenditure (% GDP)",
                             "Trade (\\% of GDP)",
                             "Non agriculture value added (\\% GDP)",
                             "GDP growth",
                             "GDP per capita (log)",
                             "Population (log)"),
          single.row = FALSE, df = FALSE,
          omit = c("country", "year"),
          add.lines = list(c("Country fixed effects", "Yes", "Yes", "Yes", "Yes"),
                           c("Yearly fixed effects", "Yes", "Yes", "Yes", "Yes")),
          omit.stat = c("rsq", "f"),
          notes = "Robust std. errors in parenthesis") #, corrected Zipf coefficient following Gabaix and Ibragimov (2011)")
```

Table \ref{ols.global} and \ref{2sls.global} report the estimation results based on the UN urban agglomeration data.
Starting with the former, we note that an agglomerative effect is only detected for the internet users and while a marginally opposing result can be seen for broadband users as indicated by the significant coefficients in columns (1) and (2).
For the telephony variables the estimation of \ref{eq3} did not yield statistically significant coefficients.
Before discussing further these results and the effects of the other control variables, we need to highlight that the main challenge of estimating equation \ref{eq3} is the potential endogenous nature of the share of internet users which might prevent us from being able to infer a truly causal effect.
Endogeneity might be an issue here as spatial structure, which is proxied by the
\color{blue}
Pareto exponent,
\color{black}
might be affected by another source, which also affects internet penetration.
For instance, economic development might affect the concentration of population in large cities and at the same time enable more people to go online.
If we do not address this issue, the coefficient for the main variable of interest will capture potential effects that internet penetration has on spatial structure, but also potential reverse causality effects that spatial structure might generate on internet penetration.
To overcome this potential problem, Table \ref{2sls.global} reports estimates of equation \ref{eq3} using two-stage least squares (2SLS) with instrumental variables (IVs).
The latter are variables which are correlated with our endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any -- of the internet and digital communications on spatial structure.
At a first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs and the other control variables are used instead of the endogenous variable to estimate eq. \ref{eq3}.
A significant effect will verify the causal impact of the internet and digital communication usage on spatial structure.

@@

stargazer(net.2sls, bb.2sls, fixed.2sls, mobile.2sls, type = "latex",
          se = list(net.2sls.rse[,"Std. Error"], bb.2sls.rse[,"Std. Error"],
                    fixed.2sls.rse[,"Std. Error"], mobile.2sls.rse[,"Std. Error"]),
          header = FALSE,
          no.space=TRUE,
          title = "2SLS estimation of equation (3)\\label{2sls.global}",
          font.size = "small",
          column.sep.width = "1pt",
          dep.var.labels=c("Pareto exponent 2000-18"),
          column.separate = c(1,1,1,1),
          covariate.labels=c("Internet users per 100 hab. (log)",
                             "Broadband users per 100 hab. (log)",
                             "Mobile phone users per 100 hab. (log)",
                             "Fixed phone users per 100 hab. (log)",
                             "Population density (log)",
                             "Government expenditure (\\% GDP)",
                             "Trade (\\% of GDP)",
                             "Non agriculture value added (\\% GDP)",
                             "GDP growth",
                             "GDP per capita (log)",
                             "Population (log)"),
          single.row = FALSE, df = FALSE,
          omit = c("country", "year"),
          omit.stat = c("rsq", "f"),
          add.lines = list(c(rownames(summ.net.2sls$diagnostics)[1], # use rownames as line name. in this case Weak insterumetns
                             round(summ.net.2sls$diagnostics[1, "statistic"], 2),
                             round(summ.bb.2sls$diagnostics[1, "statistic"], 2),
                             round(summ.fixed.2sls$diagnostics[1, "statistic"], 2),
                             round(summ.mobile.2sls$diagnostics[1, "statistic"], 2)),
                           c(rownames(summ.net.2sls$diagnostics)[2],
                             round(summ.net.2sls$diagnostics[2, "statistic"], 2),
                             round(summ.bb.2sls$diagnostics[2, "statistic"], 2),
                             round(summ.fixed.2sls$diagnostics[2, "statistic"], 2),
                             round(summ.mobile.2sls$diagnostics[2, "statistic"], 2)),
                           c("P-value",
                             round(summ.net.2sls$diagnostics[2, "p-value"], 2),
                             round(summ.bb.2sls$diagnostics[2, "p-value"], 2),
                             round(summ.fixed.2sls$diagnostics[2, "p-value"], 2),
                             round(summ.mobile.2sls$diagnostics[2, "p-value"], 2)),
                           # c(rownames(summ.net.2sls$diagnostics)[3],
                           #   round(summ.net.2sls$diagnostics[3, "statistic"], 2),
                           #   round(summ.bb.2sls$diagnostics[3, "statistic"], 2),
                           #   round(summ.fixed.2sls$diagnostics[3, "statistic"], 2),
                           #   round(summ.mobile.2sls$diagnostics[3, "statistic"], 2)),
                           # c("P-value",
                           #   round(summ.net.2sls$diagnostics[3, "p-value"], 2),
                           #   round(summ.bb.2sls$diagnostics[3, "p-value"], 2),
                           #   round(summ.fixed.2sls$diagnostics[3, "p-value"], 2),
                           #   round(summ.mobile.2sls$diagnostics[3, "p-value"], 2)),
                           c("Country fixed effects", "Yes", "Yes", "Yes", "Yes"),
                           c("Yearly fixed effects", "Yes", "Yes", "Yes", "Yes")),
          notes = c("Robust std. errors in parenthesis",
                    "IV: Female participation in labour force",
                    "First stage regressions can be found in S2"))
```

The main challenge for such an exercise is to find a valid IV.
In our case, the challenge is even bigger as we need to find a variable which is not only fit for purpose, but it also contain as few missing values as possible in order to retain the same number of observations as the OLS models.
This is particularly difficult for a diverse sample of countries as the one we are dealing with here.
We propose here the percentage of female participation in the labor force, which is directly related to digital infrastructure, but not to spatial structure (see Table \ref{desc.global} for the descriptive statistics).
\color{blue}
The literature provides empirical and theoretical evidence that ICT are correlated with female participation in labour markets mostly because they enable teleworking [@dettling2017; @arntz2019working; @watson2018digitalisation; @Lechman2020; @Valberg2020]
\color{black}
At the same time, female labor participation is not related with spatial structure especially given the diverse sample of countries.
<!-- Firstly, the correlation coefficient between the IV and the Zipf coefficient is very low ($-0.14$). -->
Although one might think that increased female labor participation may lead to relocation of households to large cities and therefore affect the spatial structure of cities, such a process varies a lot within our dataset.
Even if this might be true for developed countries with mature urban systems for which a location within a large urban center might provide opportunities for both male and female workers to find jobs, our dataset also includes countries, $41$ per cent of the GDP of which is attributed to primary activities (see Table \ref{desc.global}).
So, female labor participation might also be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the weak identification first stage F-test exceeds by far all the Stock-Yogo critical values for all, but the mobile telephony regression.
Hence, we are confident to interpret the effects of internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
\color{blue}
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engage with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
\color{black}
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and telephony penetration have further enhanced agglomeration forces, at least for large urban agglomerations, while the opposite happens for fixed telephony.
These results are robust against endogeneity issues, but are limited to urban agglomeration included in our data.
Indeed, for the estimation of the
\color{blue}
Pareto exponent
\color{black}
we only included urban agglomeration of $300,000$ inhabitants or more due to data availability.
Hence, the above estimations cannot verify whether such an effect is also valid for smaller cities.
Interestingly, the above results regarding the role of fixed telephony are in agreement with previous results from [@ioannides2008effect] [@tranos2020ict] which used a smaller panel of countries but without constraints regarding the city size.
However, our results do not support the findings of [@tranos2020ict] regarding internet usage.

\color{blue}
It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
does not make up for physical distance; individuals nearer one another
communicate more.  
\color{black}

In order to overcome the city size limitation, the next section presents two case studies, for which we have obtained much more granular data and therefore we are in a position to test the effect of the internet on the tail of the urban population distribution for these countries.

@@

stargazer(as.data.frame(data.for.decr),
          type="latex",
          summary = T,
          summary.stat = c("n", "min", "mean", "sd", "max"),
          header = FALSE,
          no.space=TRUE,
          title = "Descriptive statistics for the USA model\\label{desc.usa}",
          font.size = "small",
          single.row = FALSE,
          column.sep.width = "1pt",
          covariate.labels=c("difference in ranks, 2013-18",
                             "households w. internet 2013, \\%",
                             "population 2013",
                             "\\% of unemployment 2013",
                             "\\% of white population 2013",
                             "income 2013",
                             "population density 2013",
                             "employment in service 2011, \\%",
                             "commute in minutes 2013",
                             "pop. above 25 w. Bachelor's \\\\
                             degree 2005, \\%",
                             "commute in minutes, 2005"))
```

The LHS variable we use here is the outcome of equation \ref{eq5}.
\color{blue}
We use this variable to estimate \ref{eq4} with OLS.
Additionally, we employ a normalized form of the dependent variable, so as that
variable is bounded between $0$ and $1$ and with a mean of  $0.5$^[$$Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.$$].
This transformation enables us to estimate \ref{eq4}
with a quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a difference between
two count variables, which may also assume negative values.
\color{blue}
The results of the different estimations remain qualitatively the same regardless of the form of the LHS variable we use.
Table \ref{ols.usa} reports the results of the estimation of equation \ref{eq4} using OLS.
The sign and significance of the main variable of interest -- percentage of population with computer and broadband connection -- verifies our global level results.
That is an increase in internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that are OLS and GLM.
In addition, Table \ref{ols.usa} also reports the estimates of interaction effects between the share of population with broadband connection with population and population density.
Although the interaction terms are not significant, the main effect remains qualitatively unchanged.

@@

data.for.decr <- data %>%
  dplyr::select(r_diff, download2011, pop2011, n.tests2011,  
         unemployment.share, british.pop.share, pop.dens,
         work.home.share, services.share, uni.freq) %>%
  mutate(n.tests2011perhab = n.tests2011/pop2011) %>%
  dplyr::select(-n.tests2011) %>%
  dplyr::select(r_diff, download2011, pop2011, n.tests2011perhab,  # re-redring
         unemployment.share, british.pop.share, pop.dens,
         work.home.share, services.share, uni.freq)
